A comparison of design-based and model-based approaches for finite population spatial sampling and inference.

### 2022-04-19 inst/manuscript/manuscript.tex
more minor updates

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference rest on fundamentally different foundations. In the
    design-based approach, inference relies on random sampling. In the
    model-based approach, inference relies on distributional assumptions.
    We compare the approaches for finite population spatial data.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulated and real
    data from the United States Environmental Protection Agency's 2012
    National Lakes Assessment. A variety of sample sizes, location
    layouts, dependence structures, and response types are considered. The
    population mean is the parameter of interest and performance is
    measured using statistics like bias, squared error, and interval
    coverage.
  \item
    When studying the simulated and real data, we found that regardless of
    the strength of spatial dependence in the data, the generalized random
    tessellation stratified (GRTS) algorithm, which explicitly
    incorporates spatial locations into sampling, tends to outperform the
    simple random sampling (SRS) algorithm, which does not explicitly
    incorporate spatial locations into sampling. We also found that
    model-based inference tends to outperform design-based inference, even
    for skewed data where the model-based distributional assumptions are
    violated. The performance gap between design-based inference and
    model-based inference is small when GRTS samples are used but large
    when SRS samples are used, suggesting that the sampling choice
    (whether to use GRTS or SRS) is most important when performing
    design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial sampling and
    inference that practitioners must consider when choosing between them.
    We provide relevant background contextualizing each approach and study
    their properties in a variety of scenarios, making recommendations for
    use based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

When data cannot be collected for all units in a population (i.e.,
population units), data are collected on a subset of the population
units -- this subset is called a sample. There are two general
approaches for using samples to make frequentist statistical inferences
about a population: design-based and model-based. In the design-based
approach, inference relies on randomly assigning some population units
to be in the sample (random sampling). Alternatively, in the model-based
approach, inference relies on distributional assumptions about the
underlying data-generating stochastic process (superpopulation). Each
paradigm has a deep historical context (Sterba, 2009) and its own set of
benefits and drawbacks (Brus and De Gruijter, 1997; Hansen et al.,
1983). In this manuscript, we compare design-based and model-based
approaches for finite population spatial sampling and inference.

@@

The rest of the manuscript is organized as follows. In Section
\ref{sec:background}, we introduce and provide relevant background for
design-based and model-based approaches to finite population spatial
sampling and inference. In Section \ref{sec:mm}, we describe how we
intend to compare performance of the approaches using simulated and real
data. In Section \ref{sec:results}, we present analysis reslts for the
simulated data and real data from the United States Environmental
Protection Agency's 2012 National Lakes Assessment (USEPA, 2012). And in
Section \ref{sec:discussion}, we end with a discussion and provide
directions for future research.

@@

Fundamentally, the design-based approach combines the randomness of the
sampling design with the data collected via the sample to justify the
estimation and uncertainty quantification of fixed, unknown parameters
of a population (e.g., a population mean). Treating the data as fixed
and incorporating randomness through the sampling design yields
estimators having very few other assumptions. Confidence intervals for
these types of estimators are typically derived using limiting arguments
that incorporate all possible samples. Sample means, for example, are
asymptotically normal (Gaussian) by the Central Limit Theorem (under
some assumptions). If we repeatedly select samples from the population,
then 95\% of all 95\% confidence intervals constructed from a procedure
with appropriate coverage will contain the true fixed population mean.
Särndal et al. (2003) and Lohr (2009) provide thorough reviews of the
design-based approach.

@@

It is also important to quantify the uncertainty in \(\hat{\tau}_{ht}\).
Horvitz and Thompson (1952) and Sen (1953) provide variance estimators
for \(\hat{\tau}_{ht}\), but these estimators have two drawbacks. First,
they rely on calculating \(\pi_{ij}\), the probability that population
unit \(i\) and population unit \(j\) are both in the sample -- this
quantity can be challenging if not impossible to calculate analytically
for GRTS samples. Second, these estimators tend to ignore the spatial
locations of the population units. To address these two drawbacks
simultaneously, Stevens and Olsen (2003) proposed the local neighborhood
variance estimator. The local neighborhood variance estimator does not
rely on \(\pi_{ij}\) and estimates the variance of \(\hat{\tau}\)
conditional on the random properties of the GRTS sample -- the idea
being that this conditioning should yield a more precise estimate of
\(\hat{\tau}\). They show that the contribution from each sample unit
(population unit in the sample) to the overall variance is dominated by
local variation. Thus the local neighborhood variance estimator is a
weighted sum of variance estimates from each sample unit's local
neighborhood. These local neighborhoods contain the sample unit itself
and its three nearest neighbors among all other sample units. For more
details, see Stevens and Olsen (2003).

@@

Finite population block kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FPBK
next -- see Ver Hoef (2008) for technical details and see Higham et al.
(2021) for an extension to cases of imperfect detection among population
units. Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal \(1 / N\)). Denoting quantities that are part of the sampled
population units with a subscript \emph{s} and quantities that are part
of the unsampled population units with a subscript \emph{u}, let

@@

FPBK assumes \(\bm{\delta}\) in Equation\(~\)\eqref{equation:Zmarginal}
has mean-zero and a spatial dependence structure that can be modeled
using a covariance function. This covariance function is commonly
assumed to be non-negative, second-order stationary (depending only on
the separation vector (e.g., distance) between population units),
isotropic (independent of direction), and decays with distance between
population units (Cressie, 1993). Henceforth, it is implied that we have
made these same assumptions regarding \(\bm{\delta}\). Chiles and
Delfiner (1999), pp.~80-93 discuss covariance functions that are not
second-order stationary, not isotropic, or not either. A variety of
flexible covariance functions can be used to model \(\bm{\delta}\)
(Cressie, 1993); one example is the exponential covariance function
(Cressie (1993) provides a thorough list of spatial covariance
functions). The \(i,j\)th element of the exponential covariance matrix,
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\), is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is the variance parameter that
quantifies the spatially dependent (correlated) variability,
\(\sigma^2_{2}\) is the variance parameter the quantifies that spatially
independent (not correlated) variability, \(\phi\) is the distance
parameter that measures the distance-decay rate of the covariance, and
\(h_{i,j}\) is the Euclidean distance between population units \(i\) and
\(j\). In geostatistical literature, \(\sigma^2_{1}\) is called the
partial sill, \(\sigma^2_{2}\) is called the nugget, and \(\phi\) is
called the range. We denote \(\bm{\theta}\) as the vector of covariance
parameters that composes \(\bm{\delta}\). In
Equation\(~\)\ref{equation:expcov},
\(\bm{\theta} = \{\sigma^2_1, \sigma^2_2, \phi \}\).

The parameters in Equation\(~\)\ref{equation:Zmarginal} can be estimated
using a variety of techniques, but we focus on using restricted maximum
likelihood (REML) (Harville, 1977; Patterson and Thompson, 1971;
Wolfinger et al., 1994). REML is preferred over maximum likelihood (ML)
because ML estimates can be badly biased for small sample sizes, due to
the fact that ML makes no adjustment for the simultaneous estimation of
\(\bm{\beta}\) and \(\bm{\delta}\) (Patterson and Thompson, 1971). Minus
twice the REML log-likelihood of the sampled sites is given by
\begin{equation}\label{equation:reml}
  \ln|\bm{\Sigma}| + (\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}})^T \bm{\Sigma}_{ss}^{-1}(\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}}) + \ln|\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s| + (n - p) \ln(2 \pi) ,
\end{equation} where
\(\bm{\tilde{\beta}} = (\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s)^{-1} \bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{z}_s\)
and \(\bm{\Sigma}_{ss}\) is the covariance matrix of the sampled sites.
Minimizing Equation\(~\)\ref{equation:reml} yields
\(\bm{\hat{\theta}}_{reml}\), the REML estimates of \(\bm{\delta}\).
Then \(\bm{\hat{\beta}}_{reml}\), the REML estimate of \(\bm{\beta}\),
is given by
\((\bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{X})^{-1} \bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{z}_s\),
where \(\bm{\hat{\Sigma}}_{ss}\) is \(\bm{\Sigma}_{ss}\) evaluated at
\(\bm{\hat{\theta}}_{reml}\).

@@

In this section we describe how we used simulated and real data to
investigate performance between simple random sampling (SRS) and GRTS
sampling as well as performance between design-based (DB) and
model-based (MB) inference. In SRS and GRTS sampling, all population
units had equal inclusion probabilities and were selected without
replacement. The important distinction between SRS and GRTS is that SRS
ignores spatial locations while sampling but GRTS explicitly
incorporates them. Together, the two sampling plans (SRS and GRTS)
combined with the two inference approaches (DB and MB) yielded four
sampling-inference combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB.
For SRS-DB, the Horvitz-Thompson estimator \eqref{eq:ht} was used to
estimate means and the commonly-used SRS variance formula (Lohr, 2009;
Särndal et al., 2003) was used to estimate the variance. This variance
formula is given by \begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation} where \(z_i\) is the \(i\)th response value, \(\bar{z}\)
is the mean of all \(z_i\), \(n\) is the sample size, \(N\) is the
population size, and \(f = (1 - n / N)\) (\(f\) is often called the
finite population correction factor). For GRTS-DB, the Horvitz-Thompson
esetimator was used to estimate means and the local neighborhood
variance was used to estimate variances. For SRS-MB and GRTS-MB, FPBK
was used to estimate means and variances using restricted maximum
likelihood.

We used simulated data to compare the sampling-inference combinations
across many realized populations from the same data-generating
stochastic process. With the simulated data, we were in control of the
data-generating stochastic process and the random sampling process. We
then used real data from the 2012 National Lakes Assessment to compare
the sampling-inference combinations within a single realized population
(which is typically the case in reality). With the real data, we were in
control of only the random sampling process.

@@

\hypertarget{sec:mm_app}{%
\subsection{National Lakes Assessment (Real) Data}\label{sec:mm_app}}

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the contiguous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality.
We focus on analyzing zooplankton multi-metric indices (ZMMI) and
mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA.
For ZMMI, data were collected at 1035 unique lakes. At less than 10\% of
lakes, two ZMMI replicates were collected. These were averaged for the
purposes of our study so that each lake had one measurement for ZMMI.
For Hg ppb, data were collected at 995 unique lakes (there were no
replicates). The ZMMI and Hg ppb data are shown as spatial maps and as
histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest
near the coasts, lowest in the Central United States, are relatively
symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest
in the Northeastern United States, lowest elsewhere, are skewed, and
have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate
spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram
quantifies the halved average squared differences (semivariance) of
responses whose separation (distance) falls within some distance class.
The spatial semivariance is closely related to the spatial covariance,
and spatial semivariograms are often used to gauge the strength of
spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately
strong spatial dependence (Figure \ref{fig:zmmi}), as the semivariance
increases steadily with distance (meaning that observations nearby one
another tend to be more similar than observations far apart from one
another).

\begin{figure}
\centering
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_map.jpeg}
  \caption{Spatial map of the ZMMI population.}
  \label{fig:zmmi_map}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_map.jpeg}
  \caption{Spatial map of mercury (Hg ppb) population.}
  \label{fig:mercury_map}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_hist.jpeg}
  \caption{Histogram of the ZMMI population.}
  \label{fig:zmmi_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_hist.jpeg}
  \caption{Histogram of the mercury (Hg ppb) population.}
  \label{fig:mercury_hist}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_sv.jpeg}
  \caption{Semivariogram of the ZMMI population.}
  \label{fig:zmmi_sv_plot}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_sv.jpeg}
  \caption{Semivariogram of the mercury (Hg ppb) population.}
  \label{fig:mercury_sv_plot}
\end{subfigure}
\caption{Exploratory graphics representing populations for the zooplankton multi-metric indices (ZMMI) and mercury concentration in parts per billion (Hg ppb) in the 2012 National Lakes Assessment (NLA) data.}
\label{fig:zmmi}
\end{figure}

We studied performance of the four sampling-inference combinations by
selecting 2000 random IRS and GRTS samples of size \(n = 50\),
\(n = 100\), and \(n = 200\) from the realized ZMMI and Hg ppb
populations and then analyzing the samples using MB and DB inference. In
total, there were six separate scenarios (two responses crossed with
three sample sizes). We used the same evaluation metrics as for the
simulated data: mean bias, RMSE, and interval coverage. Mean bias is
taken as the average deviation between each sample's estimated (or
predicted) mean (\(\hat{\mu}_i\)) and the population mean (\(\mu\)) (of
ZMMI or Hg ppb): \(\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)\),
where \(i\) indexes simulation trials. Root-mean-squared error is taken
as the square root of the average squared deviation between each
sample's estimated (or predicted) mean and its population mean:
\(\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)^2}\).
Interval coverage is taken as the proportion of simulation trials where
the population mean is contained in its 95\% confidence (or prediction)
interval. These intervals are constructed using the normal distribution.

@@

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative RMSE in the simulation study for the four sampling-inference combinations and three sample sizes in the random location layout. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black lines separate the sample sizes.}
  \label{fig:rmspe_eff}
\end{figure}

95\% interval coverage for each of the four sampling-inference
combinations in the random location layout is shown in Fig.
\ref{fig:figconf}. Within each simulation scenario, all
sampling-inference combinations tend to have fairly similar interval
coverage, though when \(n = 50\) or \(n = 100\), GRTS-DB coverage is
usually a few percentage points lower than the other combinations, which
suggests that the local neighborhood variance estimate may be slightly
too small for small \(n\). Coverage in the normal response scenarios is
usually near 95\%, while coverage in the skewed response scenarios
usually varies from 90\% to 95\% but increases with the sample size. At
a sample size of 200, all four sampling-inference combinations have
approximately 95\% interval coverage in both response scenarios for all
dependent random error proportions. These interval coverage conclusions
are similar to those observed in the grid location layout, so we omit a
figure and discussion regarding the grid location layout here. Tables
for interval coverage in all 36 simulation scenarios are provided in the
supporting information.

>>>>

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference rest on fundamentally different foundations. In the
    design-based approach, inference relies on random sampling. In the
    model-based approach, inference relies on distributional assumptions.
    We compare the approaches in a finite population spatial context.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulated and real
    data from the United States Environmental Protection Agency's 2012
    National Lakes Assessment. A variety of sample sizes, location
    layouts, dependence structures, and response types are considered. The
    population mean is the parameter of interest and performance is
    measured using statistics like bias, squared error, and interval
    coverage.
  \item
    When studying the simulated and real data, we found that regardless of
    the strength of spatial dependence in the data, the generalized random
    tessellation stratified (GRTS) algorithm, which explicitly
    incorporates spatial locations into sampling, tends to outperform the
    simple random sampling (SRS) algorithm, which does not explicitly
    incorporate spatial locations into sampling. We also found that
    model-based inference tends to outperform design-based inference, even
    for skewed data where the model-based distributional assumptions are
    violated. The performance gap between design-based inference and
    model-based inference is small when GRTS samples are used but large
    when SRS samples are used, suggesting that the sampling choice
    (whether to use GRTS or SRS) is most important when performing
    design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial sampling and
    inference that practitioners must consider when choosing between them.
    We provide relevant background contextualizing each approach and study
    their properties in a variety of scenarios, making recommendations for
    use based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

When data cannot be collected for all units in a population (population
units), data are collected on a subset of the population units -- this
subset is called a sample. There are two general approaches for using
samples to make frequentist statistical inferences about a population:
design-based and model-based. In the design-based approach, inference
relies on randomly assigning some population units to be in the sample
(random sampling). Alternatively, in the model-based approach, inference
relies on distributional assumptions about the underlying
data-generating stochastic process (superpopulation). Each paradigm has
a deep historical context (Sterba, 2009) and its own set of benefits and
drawbacks (Brus and De Gruijter, 1997; Hansen et al., 1983). In this
manuscript, we compare design-based and model-based approaches for
finite population spatial sampling and inference.

@@

The rest of the manuscript is organized as follows. In Section
\ref{sec:background}, we introduce and provide relevant background for
design-based and model-based approaches to finite population spatial
sampling and inference. In Section \ref{sec:mm}, we describe how we
intend to compare performance of the approaches using simulated and real
data. The real data is from the United States Environmental Protection
Agency's 2012 National Lakes Assessment (NLA) (USEPA, 2012). In Section
\ref{sec:results}, we present analysis results for the simulated data
and NLA data. And in Section \ref{sec:discussion}, we end with a
discussion and provide directions for future research.

@@

Fundamentally, the design-based approach combines the randomness of the
sampling design with the data collected via the sample to justify the
estimation and uncertainty quantification of fixed, unknown parameters
of a population (e.g., a population mean). Treating the data as fixed
and incorporating randomness through the sampling design yields
estimators having very few other assumptions. Confidence intervals for
these types of estimators are typically derived using limiting arguments
that incorporate all possible samples. Sample means, for example, are
asymptotically normal (Gaussian) by the central limit theorem (under
some assumptions). If we repeatedly select samples from the population,
then 95\% of all 95\% confidence intervals constructed from a procedure
with appropriate coverage will contain the true fixed population mean.
Särndal et al. (2003) and Lohr (2009) provide thorough reviews of the
design-based approach.

@@

It is also important to quantify the uncertainty in \(\hat{\tau}_{ht}\).
The Horvitz-Thompson (Horvitz and Thompson, 1952) and Sen-Yates-Grundy
(Sen, 1953; Yates and Grundy, 1953) variance estimators are often used
to estimate \(\text{Var}(\hat{\tau}_{ht})\), but these estimators have
two drawbacks. First, they rely on calculating \(\pi_{ij}\), the
probability that population unit \(i\) and population unit \(j\) are
both in the sample -- this quantity can be challenging if not impossible
to calculate analytically for GRTS samples. Second, these estimators
tend to ignore the spatial locations of the population units. To address
these two drawbacks simultaneously, Stevens and Olsen (2003) proposed
the local neighborhood variance estimator. The local neighborhood
variance estimator does not rely on \(\pi_{ij}\) and estimates the
variance of \(\hat{\tau}\) conditional on the random properties of the
GRTS sample -- the idea being that this conditioning should yield a more
precise estimate of \(\hat{\tau}\). They show that the contribution from
each sampled population unit to the overall variance is dominated by
local variation. Thus the local neighborhood variance estimator is a
weighted sum of variance estimates from each sampled population unit's
local neighborhood. These local neighborhoods contain the sampled
population unit itself and its three nearest neighbors (among all other
sampled population units). For more details, see Stevens and Olsen
(2003).

@@

Finite population block kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FPBK
next -- see Ver Hoef (2008) for technical details and see Higham, J. Ver
Hoef, et al. (2021) for an extension to cases of imperfect detection
among population units. Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal \(1 / N\)). Denoting quantities that are part of the sampled
population units with a subscript \emph{s} and quantities that are part
of the unsampled population units with a subscript \emph{u}, let

@@

FPBK assumes \(\bm{\delta}\) in Equation\(~\)\eqref{equation:Zmarginal}
has mean-zero and a spatial dependence structure that can be modeled
using a covariance function. This covariance function is commonly
assumed to be non-negative, second-order stationary (depending only on
the separation vector (e.g., distance) between population units),
isotropic (independent of direction), and decays with distance between
population units (Cressie, 1993). Henceforth, it is implied that we have
made these same assumptions regarding \(\bm{\delta}\). Chiles and
Delfiner (1999), pp.~80-93 discuss covariance functions that are not
second-order stationary, not isotropic, or not either. A variety of
flexible covariance functions can be used to model \(\bm{\delta}\)
(Cressie, 1993) -- one example is the exponential covariance function.
Cressie (1993) provides a thorough list of spatial covariance functions.
The \(i,j\)th element of the exponential covariance matrix,
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\), is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is the variance parameter that
quantifies the spatially dependent (correlated) variability,
\(\sigma^2_{2}\) is the variance parameter the quantifies that spatially
independent (not correlated) variability, \(\phi\) is the distance
parameter that measures the distance-decay rate of the covariance, and
\(h_{i,j}\) is the Euclidean distance between population units \(i\) and
\(j\). In geostatistical literature, \(\sigma^2_{1}\) is called the
partial sill, \(\sigma^2_{2}\) is called the nugget, and \(\phi\) is
called the range. We denote \(\bm{\theta}\) as the vector of covariance
parameters that composes \(\bm{\delta}\). In
Equation\(~\)\ref{equation:expcov},
\(\bm{\theta} = \{\sigma^2_1, \sigma^2_2, \phi \}\).

The parameters in Equation\(~\)\ref{equation:Zmarginal} can be estimated
using a variety of techniques, but we focus on using restricted maximum
likelihood (REML) (Harville, 1977; Patterson and Thompson, 1971;
Wolfinger et al., 1994). REML is preferred over maximum likelihood (ML)
because ML estimates can be badly biased for small sample sizes, due to
the fact that ML makes no adjustment for the simultaneous estimation of
\(\bm{\beta}\) and \(\bm{\theta}\) (Patterson and Thompson, 1971). Minus
twice the REML log-likelihood of the sampled sites is given by
\begin{equation}\label{equation:reml}
  \ln|\bm{\Sigma}| + (\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}})^T \bm{\Sigma}_{ss}^{-1}(\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}}) + \ln|\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s| + (n - p) \ln(2 \pi) ,
\end{equation} where
\(\bm{\tilde{\beta}} = (\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s)^{-1} \bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{z}_s\)
and \(\bm{\Sigma}_{ss}\) is the covariance matrix of the sampled sites.
Minimizing Equation\(~\)\ref{equation:reml} yields
\(\bm{\hat{\theta}}_{reml}\), the REML estimates of \(\bm{\theta}\).
Then \(\bm{\hat{\beta}}_{reml}\), the REML estimate of \(\bm{\beta}\),
is given by
\((\bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{X})^{-1} \bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{z}_s\),
where \(\bm{\hat{\Sigma}}_{ss}\) is \(\bm{\Sigma}_{ss}\) evaluated at
\(\bm{\hat{\theta}}_{reml}\).

@@

In this section we describe how we used simulated and real data to
investigate performance between simple random sampling (SRS) and GRTS
sampling as well as performance between design-based (DB) and
model-based (MB) inference. In SRS and GRTS sampling, all population
units had equal inclusion probabilities and were selected without
replacement. The important distinction between SRS and GRTS is that SRS
ignores spatial locations while sampling but GRTS explicitly
incorporates them. Together, the two sampling plans (SRS and GRTS)
combined with the two inference approaches (DB and MB) yielded four
sampling-inference combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB.
For SRS-DB, the Horvitz-Thompson estimator \eqref{eq:ht} was used to
estimate means and the commonly-used SRS variance formula (Lohr, 2009;
Särndal et al., 2003) was used to estimate the variance. This variance
formula is given by \begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation} where \(z_i\) is the \(i\)th response value, \(\bar{z}\)
is the mean of all \(z_i\), \(n\) is the sample size, \(N\) is the
population size, and \(f = (1 - n / N)\) (\(f\) is often called the
finite population correction factor). For GRTS-DB, the Horvitz-Thompson
estimator was used to estimate means and the local neighborhood variance
was used to estimate variances. For SRS-MB and GRTS-MB, FPBK was used to
estimate means and variances using restricted maximum likelihood. SRS,
GRTS sampling, and design-based inference were implemented using the
\texttt{spsurvey} \textbf{\textsf{R}} package (Dumelle et al., 2022).
FPBK was implemented using the \texttt{sptotal} \textbf{\textsf{R}}
package (Higham, Ver Hoef, et al., 2021).

We used simulated data to compare the sampling-inference combinations
across many realized populations from the same data-generating
stochastic process. We then used real data from the 2012 National Lakes
Assessment to compare the sampling-inference combinations within a
single realized population (which is typically the case in reality).
With the simulated data, we were in control of the data-generating
stochastic process and the random sampling process. With the real data,
we were in control of only the random sampling process.

@@

\hypertarget{sec:mm_app}{%
\subsection{National Lakes Assessment Data}\label{sec:mm_app}}

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the conterminous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality.
We focus on analyzing zooplankton multi-metric indices (ZMMI) and
mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA.
For ZMMI, data were collected at 1035 unique lakes. At less than 10\% of
lakes, two ZMMI replicates were collected. These were averaged for the
purposes of our study so that each lake had one measurement for ZMMI.
For Hg ppb, data were collected at 995 unique lakes (there were no
replicates). The ZMMI and Hg ppb data are shown as spatial maps and as
histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest
near the coasts, lowest in the Central United States, are relatively
symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest
in the Northeastern United States, lowest elsewhere, are skewed, and
have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate
spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram
quantifies the halved average squared differences (semivariance) of
responses whose separation (distance) falls within some distance class.
The spatial semivariance is closely related to the spatial covariance,
and spatial semivariograms are often used to gauge the strength of
spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately
strong spatial dependence (Figure \ref{fig:zmmi}), as the semivariance
increases steadily with distance (meaning that observations nearby one
another tend to be more similar than observations far apart from one
another).

\begin{figure}
\centering
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_map.jpeg}
  \caption{Spatial map of the ZMMI population.}
  \label{fig:zmmi_map}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_map.jpeg}
  \caption{Spatial map of the Hg ppb population.}
  \label{fig:mercury_map}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_hist.jpeg}
  \caption{Histogram of the ZMMI population.}
  \label{fig:zmmi_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_hist.jpeg}
  \caption{Histogram of the Hg ppb population.}
  \label{fig:mercury_hist}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_sv.jpeg}
  \caption{Semivariogram of the ZMMI population.}
  \label{fig:zmmi_sv_plot}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_sv.jpeg}
  \caption{Semivariogram of the Hg ppb population.}
  \label{fig:mercury_sv_plot}
\end{subfigure}
\caption{Exploratory graphics representing populations for the zooplankton multi-metric indices (ZMMI) and mercury concentration in parts per billion (Hg ppb) in the 2012 National Lakes Assessment (NLA) data.}
\label{fig:zmmi}
\end{figure}

We studied performance of the four sampling-inference combinations by
selecting 2000 random IRS and GRTS samples of size \(n = 50\),
\(n = 100\), and \(n = 200\) from the realized ZMMI and Hg ppb
populations and then analyzing the samples using MB and DB inference. In
total, there were six separate scenarios (two responses crossed with
three sample sizes). For each sample from each scenario, parameters were
estimated using the same approaches as described for simulated data. We
also used the same evaluation metrics as for the simulated data: mean
bias, RMSE, and interval coverage. Mean bias is taken as the average
deviation between each sample's estimated (or predicted) mean
(\(\hat{\mu}_i\)) and the population mean (\(\mu\)) (of ZMMI or Hg ppb):
\(\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)\), where \(i\)
indexes simulation trials. Root-mean-squared error is taken as the
square root of the average squared deviation between each sample's
estimated (or predicted) mean and its population mean:
\(\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)^2}\).
Interval coverage is taken as the proportion of simulation trials where
the population mean is contained in its 95\% confidence (or prediction)
interval. These intervals are constructed using the normal distribution.

@@

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Simulated data relative RMSE for the four sampling-inference combinations and three sample sizes in the random location layout. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black lines separate the sample sizes.}
  \label{fig:rmspe_eff}
\end{figure}

95\% interval coverage for each of the four sampling-inference
combinations in the random location layout is shown in Fig.
\ref{fig:figconf}. Within each simulation scenario, all
sampling-inference combinations tend to have fairly similar interval
coverage, though when \(n = 50\) or \(n = 100\), GRTS-DB coverage is
usually a few percentage points lower than the other combinations, which
suggests that the local neighborhood variance estimate may be slightly
too small for small \(n\). Coverage in the normal response scenarios is
usually near 95\%, while coverage in the skewed response scenarios
usually varies from 90\% to 95\% but increases with the sample size. At
a sample size of 200, all four sampling-inference combinations have
approximately 95\% interval coverage in both response scenarios for all
DRE proportions. These interval coverage conclusions are similar to
those observed in the grid location layout, so we omit a figure and
discussion regarding the grid location layout here. Tables for interval
coverage in all 36 simulation scenarios are provided in the supporting
information.
